Spatial Coordination Of Stomatal Patterning Between Leaf Surfaces In Amphistomatous *Arabidopsis thaliana* Incurs No Photosynthetic Advantage

### 2023-10-16 ms/ms.Rmd
updated caption in fig:density and removed extra model description from ms

  It consists of two paragraphs.
keywords:
  - amphistomy
  - Arabidopsis thaliana
  - CO$_2$ diffusion
  - finite element method
  - optimality
  - photosynthesis
  - stomata
fontsize: 12pt
#spacing: halfline # could also be oneline
#classoptions:
#  - endnotes
bibliography: stomata-spacing.bib
#csl: mycitationstyle.csl
link-citations: yes
output: 
  rticles::oup_article:
    latex_engine: xelatex
#urlcolor: orange
#linkcolor: green
#citecolor: red
header-includes:
 - \usepackage[nomarkers,tablesfirst]{endfloat} # For figures and tables at end
 - \usepackage{lineno} # For line numbering
 - \linenumbers # For line numbering
 - \usepackage{hyperref} # For autoref
 - \renewcommand{\figureautorefname}{Fig.}
---

@@

We modeled photosynthesis CO$_2$ assimilation rate using a spatially-explicit two-dimensional reaction diffusion model using a porous medium approximation [@parkhurst_intercellular_1994] using the finite element method (FEM) following @earles_excess_2017. Consider a two-dimensional leaf where stomata occur on each surface in a regular sequence with interstomatal distance $U$. The main outcome we assessed is the advantage of offsetting the position of stomata on each surface compared to have stomata on the same $x$ position on each surface. With these assumptions, by symmetry, we only need to model two stomata, one abaxial and one adaxial, from $x = 0$ to $x = U/2$ and from the adaxial surface at $y = 0$ to the abaxial surface at $y = L$, the leaf thickness. We arbitrarily set the adaxial stomate at $x = 0$ and toggled the abaxial stomata position between $x = U/2$ (offset) or $x = 0$ (below adaxial stomate). The 'coordination advantage' of offsett stomatal position on each surface is the photosynthetic rate of the leaf with offset stomata compared to that with stomata aligned in the same $x$ position:

@@

We modeled the coordination advantage over a range of leaf thicknesses, stomatal densities, photosynthetic capacities, and light environments to understand when offsetting stomatal position on each surface might deliver a significant photosynthetic advantage (\autoref{tab:model_var}).

@@

knitr::kable(
  model_var, 
  caption = model_var_caption,
  label = "model_var",
  booktabs = TRUE, format = "latex", escape = FALSE
) #|>
#   kableExtra::kable_styling()
```

WORKING HERE - ???

### Light propogation model

Irradiance at depth $y$ in a leaf with thickness $L$ is modeled following Lloyd *et al.* (1992):

\begin{equation}\label{eq:light_propogation}
  I(y) = 1.1 I_0 e ^ {-2.4 y / L}
\end{equation}

where $I_0$ is photosynthetically active irradiance incident on the adaxial leaf surface.

### Biochemical model

All parameter symbols, units, descriptions, and values are described in Table X below. Following Gutschick (1984), we modeled photosynthetic rate per unit chlorophyll $A_\text{chl}$ then calculated the volumetric photosynthetic rate $A_\text{volume}$ by multiplying by the chlorophyll concentration $Q_\text{chl}$:

$$A_\text{volume} = A_\text{chl} Q_\text{chl}$$
A description of the model is given on page 553-556 of Gutschick (1984). The R code below is how we implemented the model.

* `C_m` is a vector CO$_2$ concentrations in $\text{mmol m}^{-3}$ at different positions within the leaf
* `I_m` is a vector of irradiances in $\mu \text{mol m}^{-2} \text{s}^{-1}$ at different positions within the leaf (same order as `C_m`).
* `pars` is a list of parameters

```{r, echo = TRUE, eval = FALSE}

pars = list(

    P = set_units(101.325, kPa), # air pressure at sea level
    temp = set_units(298.15, K), # assume constant temperature
    R_gas = set_units(8.314, J/K/mol), # ideal gas constant

    # Calculations for C_a and O ued below
    # 21% O2
    # set_units(0.21 * P / (R_gas * temp), mol/m^3)
    # 415 ppm
    # set_units((415/1e6) * P / (R_gas * temp), mmol/m^3)

    # Environmental
    C_a = set_units(16.96367, mmol/m^3),
    O = set_units(8.584027, mol/m^3),
    PAR = set_units(1000, umol/m^2/s),

    # Biochemical
    E_t = set_units(0.01, mol/mol),
    eta_t = set_units(0.59, 1),
    k_c = set_units(20, mol/mol/s),
    k_o = set_units(4.2, mol/mol/s),
    K_c = set_units(0.0184, mol/m^3),
    K_o = set_units(13.2, mol/m^3),
    Q_chl = set_units(3.3, mol/m^3),
    R_p = set_units(0.273, mol/mol),

    # Diffusivity of CO2 in leaf airspace. From Gutschick 1984. Should be updated and include temperature dependence, effect of cell packing, tortuosity, etc.
    D_mc = set_units(7e-7, m^2/s)

)

# strip units to speed up calculation
upars = purrr::map(pars, ~ {if(inherits(.x, "units")) {drop_units(.x)}})

# multiply/divide by 1e3 because C_m is mmol and parameters are in mol
k_vc = upars[["k_c"]] / (1 + (1e3 * upars[["K_c"]] / C_m) *
                           (1 + upars[["O"]] / upars[["K_o"]])) # [1/s]
# Given light_propagation assumptions, k = 2.4/L assuming no scattering
k_i = 2.4 / upars[["leaf_thickness"]] # [1/um] # I'm not sure sure this is right
b_x = k_i * I_m # [mol/m^3/s]
a_x = b_x / upars[["Q_chl"]] # [1/s]
j = 0.5 * a_x * upars[["eta_t"]] # [1/s] # eqn 9
phi = upars[["k_o"]] / upars[["k_c"]] * (upars[["O"]] / upars[["K_o"]]) /
  (1e-3 * C_m / upars[["K_c"]]) # [1]
v_cj = j / (4 + 4 * phi) # [1/s]
v_cp = k_vc * upars[["R_p"]] # [1/s]
v_cr = apply(cbind(v_cj, v_cp), 1, min) # [1/s]
v_c = apply(cbind(k_vc * upars[["E_t"]], v_cr), 1, min) # [1/s]
v_o = phi * v_c # [1/s]
A_chl = v_c - 0.5 * v_o # [1/s] # i.e mol CO2 / mol Chl / s
A_volume = A_chl * upars[["Q_chl"]] # [mol/m^3/s]

@@

To model photosynthesis and CO$_2$ transport within a two-dimensional cross section of the leaf, we built a grid of nodes of dimensions leaf thickness by half the interstomatal distance. Nodes represent the leaf mesophyll where CO$_2$ diffusion and assimilation occur.

**Table X** Glossary of mathematical symbols. The columns indicate the mathematical Symbol used in the paper, the associated symbol used in R scripts, scientific Units, and a verbal Description.

 Symbol              | Value(s) | Units            | Description
---------|--------|--------------------------|----------------------------------
Biochemical Parameters |||
$A_\text{volume}$    | NA    | $mol CO_2/m^3/s$    | volumetric assimilation rate
$A_\text{chl}$       | NA    | $mol CO2/mol Chl/s$ | assimilation rate per mol chlorophyll
$E_t$                | 0.01  | $mol / mol Chl$     | rubisco octamer concentration per chlorophyll  
$eta_t$              | 0.59  | unitless            | quantum efficiency of photoexcitation transfer to reaction centers
$k_c$                | 20    | $mol CO2/mol Rubisco/s$ at 25 C | maximal carboxylation velocity of Rubisco
$k_o$                | 4.2   | $mol O2/mol Rubisco/s$ at 25 C | maximal oxygenation velocity of Rubisco
$K_c$                | 0.0184 | $mol/m^3$ at 25 C  | Michaelis constant for CO2
$K_o$                | 13.2  | $mol/m^3$ at 25 C   | Michaelis constant for O2
$J_\text{max}$       | 0.253 | $mol e\text{-}/mol Chl/s$ | maximal electron transport rate per mol Chl
$Q_\text{chl}$       | 3.3   | $mol/m^3$           | Chl volume concentration
$v_c$                | NA    | $mol CO2/mol Chl/s$ | carboxylation rate
$v_o$                | NA    | $mol O2/mol Chl/s$  | oxygenation rate
$k_\text{vc}$        | NA    | $mol CO2/mol Rubisco/mol Chl$ | carboxylation velocity per Rubisco octamer at ambient CO2
$R_p$                | 0.273 | $mol RuBP/mol Chl$  | RuBP pool size
$k_i$                | 2.4   | $1/m^2$ or $1/L$    | light attenuation coefficient
$D_\text{mc}$        | 7e-7  | $m^2/s$             | diffusivity of CO$_2$ in mesophyll airspace
Environmental Parameters |||
$P$                  | 101.325 | $kPa$             | air pressure at sea level
$temp$               | 298.15 | $K$                | temperature
$R_\text{gas}$       | 8.314 | $J/K/mol$           | ideal gas constant
$C_a$                | 16.96367 | $mmol CO_2/m^3$  | atmospheric CO$_2$ concentration
$O$                  | 8.584027 | $mol O_2/m^3$    | atmospheric $O_2$ concentration
$PAR$                | 1000  | $umol photons/m^2/s$ | photosynthetically active radiation
Variables |||
$U_s$                | .05, .10, .15... | $mm$     | interstomatal distance
$T_l$                | .1, .2, .3... | $mm$        | leaf thickness
$g_\text{sc}$        | 0.2, 0.3, 0.4... | $mol/m^2/s$ | stomatal conductance

## Two-dimensional model

We used the `steady.2D()` function in the *R* package **rootSolve** version `r packageVersion("rootSolve")` [@soetaert_practical_2009].

```{r, echo = TRUE, eval=FALSE}

diffusion2D <- function (time, state, pars, stomata_offset) {

  n_row = pars[["n_row"]]
  n_col = pars[["n_col"]]

  n_node = n_row * n_col

  node_length_m = node_length * 1e-6 # node_length in [m]

  # matrix of C_m values
  C_m_mat = matrix(nrow = n_row, ncol = n_col, state)

  # Photosynthetic demand
  ## Light matrix
  I_mat = seq(0, by = pars[["node_length"]], length.out = n_row) |>
    light_propogation(pars[["leaf_thickness"]], pars[["PAR"]]) |>
    matrix(n_row, ncol = n_col)

  ## Biochemical model
  # Units have to be correct for this to work, but unitless = TRUE should be faster
  A = biochemical_model(C_m_mat, I_mat, pars = pars, unitless = TRUE)

  ## Photosynthesis matrix
  ## multiply by 1000 to convert from mol / m^3 / s to mmol / m^3 / s
  A_mat = matrix(nrow = n_row, ncol = n_col, 1e3 * A)

  # CO2 diffusion
  C_a = drop_units(pars[["C_a"]])
  flux_mat = matrix(nrow = n_row, ncol = n_col, 0)
  flux_mat[1, 1] = pars[["g_sc"]] * (C_a - C_m_mat[1, 1])

  ## 1. Flux through stomata
  if (stomata_offset) {
    flux_mat[n_row, n_col] = pars[["g_sc"]] * (C_a - C_m_mat[n_row, n_col])
  } else {
    flux_mat[n_row, 1] = pars[["g_sc"]] * (C_a - C_m_mat[n_row, 1])
  }

  zero_x <- rep(0, n_row)
  zero_y <- rep(0, n_col)

  ## 2. Mesophyll flux; zero fluxes near boundaries
  flux_above = rbind(rep(0, n_col), C_m_mat[1:(n_row - 1), ] - C_m_mat[2:n_row, ])
  flux_below = rbind(C_m_mat[2:n_row, ] - C_m_mat[1:(n_row - 1), ], rep(0, n_col))
  flux_left  = cbind(rep(0, n_col), C_m_mat[, 1:(n_col - 1)] - C_m_mat[, 2:n_col])
  flux_right = cbind(C_m_mat[, 2:n_col] - C_m_mat[, 1:(n_col - 1)], rep(0, n_col))

  # g_mc [m/s] = D_mc [m^2/s] / node_length [m]
  flux_mat = flux_mat + pars[["g_mc"]] * (flux_above + flux_below + flux_left + flux_right)

    return(list(c(as.vector(flux_mat / node_length_m - A_mat))))

}

## Global Results

@@

@@

\begin{figure}[ht]
\includegraphics[width = \textwidth]{figures/density.pdf}
\caption{Stomatal density is higher in plants grown under high light conditions. Significant differences between light treatments are indicated by asterisks according to analysis of variance followed by a post-hoc tukey honest significant difference test ($\alpha$ = 0.05).}
\label{fig:density}
\end{figure}

>>>>

  It consists of two paragraphs.
keywords:
  - amphistomy
  - Arabidopsis thaliana
  - CO$_2$ diffusion
  - finite element method
  - optimality
  - photosynthesis
  - stomata
fontsize: 12pt
#spacing: halfline # could also be oneline
#classoptions:
#  - endnotes
bibliography: stomata-spacing.bib
#csl: mycitationstyle.csl
link-citations: yes
output: 
  rticles::oup_article:
    latex_engine: xelatex
#urlcolor: orange
#linkcolor: green
#citecolor: red
header-includes:
 - \usepackage[nomarkers,tablesfirst]{endfloat} # For figures and tables at end
 - \usepackage{lineno} # For line numbering
 - \linenumbers # For line numbering
 - \usepackage{hyperref} # For autoref
 - \renewcommand{\figureautorefname}{Fig.}
 - \usepackage{booktabs} # for tables
---

@@

We modeled photosynthesis CO$_2$ assimilation rate using a spatially-explicit two-dimensional reaction diffusion model using a porous medium approximation [@parkhurst_intercellular_1994] using the finite element method (FEM) following @earles_excess_2017. Consider a two-dimensional leaf where stomata occur on each surface in a regular sequence with interstomatal distance $U$. The main outcome we assessed is the advantage of offsetting the position of stomata on each surface compared to have stomata on the same $x$ position on each surface. With these assumptions, by symmetry, we only need to model two stomata, one abaxial and one adaxial, from $x = 0$ to $x = U/2$ and from the adaxial surface at $y = 0$ to the abaxial surface at $y = L$, the leaf thickness. We arbitrarily set the adaxial stomate at $x = 0$ and toggled the abaxial stomata position between $x = U/2$ (offset) or $x = 0$ (below adaxial stomate). The 'coordination advantage' of offset stomatal position on each surface is the photosynthetic rate of the leaf with offset stomata compared to that with stomata aligned in the same $x$ position:

@@

We modeled the coordination advantage over a range of leaf thicknesses, stomatal densities, photosynthetic capacities, and light environments to understand when offsetting stomatal position on each surface might deliver a significant photosynthetic advantage (\autoref{tab:model_var}). The complete model description is available in the Supporting Information.

@@

knitr::kable(
  model_var, 
  caption = model_var_caption,
  label = "model_var",
  booktabs = TRUE, format = "latex", escape = FALSE
)

@@

@@

\begin{figure}[ht]
\includegraphics[width = \textwidth]{figures/density.pdf}
\caption{Stomatal density is higher in \textit{A. thaliana} plants grown under high light conditions. We determined statistical significance between light treatments using Tukey's post-hoc tests. $^*~0.05 > P \ge 0.01; ^{**}~0.01 > P \ge 0.001; ^{***}~0.0001 > P \ge 0.0001; ^{***}~ P <0.0001$.}
\label{fig:density}
\end{figure}

@@
